LabMate: supporting types for MATLAB

### 2024-04-08 writing/IMEKOWC2024/LabMate/paper.tex
[ writing ] change style file

\documentclass{ws-procs9x6}

%%%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[colorlinks=true]{hyperref}
\usepackage{microtype}
\usepackage{proof}
\usepackage{mathtools}
% Added by Keith L 05/04/2024
% Removed by FNF 06/04/2024 (incompatible with style file)
%\usepackage[]{biblatex}
%\addbibresource{ws-procs9x6.bib}
\usepackage{graphics}
%\usepackage{caption}
\usepackage{floatrow}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\title{LabMate: supporting types for MATLAB}
\author{Conor McBride, Georgi Nakov, Fredrik Nordvall Forsberg, Alistair Forbes, Keith Lines}
\address{University of Strathclyde, National Physical Laboratory}
%TODO: Affiliations

\bodymatter

@@

The recorded output follows the input signal, superimposed on a detector with noise, offset and drift. Figure~\ref{fig:rec_out} provides an example of recorded output, using simulated data.
\begin{figure}[H]
    \centering
    \captionsetup{justification=centering}
    \includegraphics[width=0.6\linewidth]{example_output.jpg}
    \caption{Simulated example of recorded output}       
    \label{fig:rec_out}
\end{figure}

@@

Another noteworthy special case which we want to support takes $R$ and
$R$ to be types of physical dimensions, with $E(d_1, d_2)$ to be the
type of physical quantities of dimension $d_1 \cdot d_2^{-1}$.
%
This way, we can reduce dimensional consistency checking \`a la
Kennedy~\cite{kennedyUOM} also for programs involving matrices,
following the work of Hart~\cite{hart}.
%
We will see in \sref{sec:example-revisited} how this can be helpful.

@@

Coming back to the example from \sref{sec:example} again, we can give both \texttt{A} and \texttt{B} more precise types, also taking their physical dimensions into account:
\begin{verbatim}
%> A :: [ i <- 12,  j <- 5 ] (if j == 4 then Q T else Q 1)
%> B :: [ i <- 5 ] (if i == 4 then Q (ML^2T^(-4)I^(-1)) else Q (ML^2T^(-3)I^(-1)))
\end{verbatim}
This says that the entries in the last column of \texttt{A} are quantities of dimension time $\mathsf{T}$, while all other entries are dimensionless quantities. Meanwhile the entries of \texttt{B} are quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$ (i.e., units V) for entries in the first three rows and $\mathsf{ML}^{2}\mathsf{T}^{-4}\mathsf{I}^{-1}$ (i.e., units V/s) respectively. LabMate reports that the type of \texttt{C} should be
\begin{verbatim}
%< C :: [ 5 x 1 ] (Q (ML^2T^(-3)I^(-1)))
\end{verbatim}
that is a $5 \times 1$ matrix whose entries are all quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$. Another possible way to use LabMate is to instead declare that this is the expected type for \texttt{C}, and to leave the type of one of the entries in \texttt{B} as an unknown:
\begin{verbatim}
%> B :: [ Q TODO; Q TODO; ?aTy; Q TODO; Q TODO; Q TODO ]
\end{verbatim}
We can then ask LabMate to solve the \remph{meta variable} \texttt{?aTy}, to which LabMate responds that we must have \todo{\texttt{?aTy} = fill in}. This way the types of LabMate can be used to validate and experiment with the dimensional consistency of the quantities involved.

@@

Thanks to NPL colleague Nick Fletcher for his help and guidance providing the example from \sref{sec:example}. Thanks also to NPL colleagues Louise Wright and Ian Smith for reviewing this paper.

>>>>

\documentclass{IMEKO2024}

%%%% Packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage{amsmath}
\usepackage[colorlinks=true]{hyperref}
\usepackage{microtype}
\usepackage{proof}
\usepackage{mathtools}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{subcaption}
%\usepackage{multirow}
\usepackage{xkeyval}
\usepackage[font=small,justification=centering]{caption}
\captionsetup{labelsep = period}
\usepackage{floatrow}

@@

\title{LabMate: supporting types for MATLAB}
\author{Conor McBride  \affiliationmark{1}, Georgi Nakov  \affiliationmark{1}, Fredrik Nordvall Forsberg  \affiliationmark{1}, Alistair Forbes  \affiliationmark{2}, Keith Lines  \affiliationmark{2}}
\affiliation{%
\affiliationmark{1}{\fontsize{11pt}{9pt} \selectfont University of Strathclyde , Glasgow, UK}  \\
\affiliationmark{2}{\fontsize{11pt}{9pt} \selectfont National Physical Laboratory, Teddington, UK}}

\maketitle

\begin{abstract}
  TODO
\end{abstract}
\begin{keywords}
  TODO
\end{keywords}

@@

The recorded output follows the input signal, superimposed on a detector with noise, offset and drift. Figure~1%\ref{fig:rec_out}
provides an example of recorded output, using simulated data.
\begin{figure}[H]
  \begin{center}
    \includegraphics[width=0.6\linewidth]{example_output.jpg}

    \captionwide{Simulated example of recorded output.}
  \end{center}
\end{figure}

@@

Another noteworthy special case which we want to support takes $R$ and
$R$ to be types of physical dimensions, with $E(d_1, d_2)$ to be the
type of physical quantities of dimension $d_1 \cdot d_2^{-1}$.
%
This way, we can reduce dimensional consistency checking \`a la
Kennedy~\cite{kennedyUOM} also for programs involving matrices,
following the work of Hart~\cite{hart}.
%
We will see in Section~\ref{sec:example-revisited} how this can be helpful.

@@

Coming back to the example from Section~\ref{sec:example} again, we can give both \texttt{A} and \texttt{B} more precise types, also taking their physical dimensions into account:
\begin{verbatim}
%> A :: [ i <- 12,  j <- 5 ] (if j == 4 then Q T else Q 1)
%> B :: [ i <- 5 ] (if i == 4 then Q (ML^2T^(-4)I^(-1)) else Q (ML^2T^(-3)I^(-1)))
\end{verbatim}
This says that the entries in the last column of \texttt{A} are quantities of dimension time $\mathsf{T}$, while all other entries are dimensionless quantities. Meanwhile the entries of \texttt{B} are quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$ (i.e., units V) for entries in the first three rows and $\mathsf{ML}^{2}\mathsf{T}^{-4}\mathsf{I}^{-1}$ (i.e., units V/s) respectively. LabMate reports that the type of \texttt{C} should be
\begin{verbatim}
%< C :: [ 5 x 1 ] (Q (ML^2T^(-3)I^(-1)))
\end{verbatim}
that is a $5 \times 1$ matrix whose entries are all quantities of dimension $\mathsf{ML}^{2}\mathsf{T}^{-3}\mathsf{I}^{-1}$. Another possible way to use LabMate is to instead declare that this is the expected type for \texttt{C}, and to leave the type of one of the entries in \texttt{B} as an unknown:
\begin{verbatim}
%> B :: [ Q TODO; Q TODO; ?aTy; Q TODO; Q TODO; Q TODO ]
\end{verbatim}
We can then ask LabMate to solve the \remph{meta variable} \texttt{?aTy}, to which LabMate responds that we must have \todo{\texttt{?aTy} = fill in}. This way the types of LabMate can be used to validate and experiment with the dimensional consistency of the quantities involved.

@@

Thanks to NPL colleague Nick Fletcher for his help and guidance providing the example from Section \ref{sec:example}. Thanks also to NPL colleagues Louise Wright and Ian Smith for reviewing this paper.
